Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-03-21 deprecated.tex
getting close

>>>>

\subsubsection{Refinements}

It will be an undertaking, but if we want to approach this using the refinements we might get something pretty nifty.
Everything will be quite messy, but really the entire approach boils down to a series of nested inequalities in two variables.
We may need a program that can reduce symbolic equations, but this will at least get us somewhere that isn't a lot of loose approximations---initial results using numerical optimization show the region in~\ref{fig:inconsistency-farris} is conservative, and very much so.
See Fig.~\ref{fig:numerical-inconsistency-farris}.

Perhaps something along the lines of a more general results would be interesting?
Like, could we calculate the measure/area of inconsistency using cooler techniques?
Could we do this for a general topology?
Would that even be interesting since we've already found a counterexample?

It also seems that, in the $\{x, y\}=\{.25, .25\}$ example, the lower bound is tight while the upper bound is looser.
This surprises me since we replaced a lot in the likelihood, but this might indicate that the $H$ term is actually the one that varies more than the $L$ term.
I guess this is intuitive: the difference between the lower and upper bound on $L$ is $\log2\approx0.69$.
Maybe getting a tighter upper bound will be useful here?

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{ineqs-num-far-gen}
\caption{Regions of inconsistency---numerically obtained}
\label{fig:numerical-inconsistency-farris}
\end{figure}

\subsubsection{Tighter upper bound}

Call
$$
e = p(0001) + p(0100) = \frac{1+x^2-y^2-x^2y^2}{4},
$$
$$
f = p(0010) + p(1000) = \frac{1-x^2+y^2-x^2y^2}{4}.
$$
Then some algebra and standard $\log$ bounds can be used to get an upper bound for the full Farris likelihood $L^{(i)}$ as
$$
(2-a+e)\log(2-\frac{2a}{2+e}) + a\log\frac{2a}{2+e} + (2-b+f)\log(2-\frac{2b}{2+f}) + a\log\frac{2b}{2+f}.
$$
In our example of $\{x,y\}=\{.25,.25\}$ this bound is $a$ compared to $-4.25313560433$ before.

### 2017-03-21 main.tex
getting close

Our key approach will be to decompose the likelihood as
\begin{equation}
\label{eq:partial-likelihood}
L^{(i)}(x', y', w') = H_{\tau'}(x,y;x',y',w') + L^{p,(i)}(x',y',w')
\end{equation}
where the first term is \eqref{eq:information} and the second term will be a weighted sum of the $\log$s of values from Table~\ref{tab:likelihoods}, the superscript indicating which choices for ancestral state likelihood values we choose.
Define
$$
\{\hat{x}, \hat{y}, \hat{w}, \hat{i}\} = \arg\max_{\{x,y,w,i\}} \ L^{p,(i)}(x,y,w).
$$
We have upper and lower bounds for \eqref{eq:partial-likelihood} in the form
$$
H_{\tau'}(x,y; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H(x,y;x,y,y) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
$$

@@

We want to find values for $\{x,y\}$ where
$$
H_{fe}(x,y; 1, 0, 1-c-2d) + \hat{L}_{fe}(c,d) \ge H_{fa}(x,y; x, y, y) + \hat{L}_{fa}(a,b).
$$
We can cheat a bit and plot the values at which this is true.
This may give us some insight into what form these regions of inconsistency take---see Fig.~\ref{fig:inconsistency-farris}.

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{ineqs-max-far-gen}
\caption{Regions of inconsistency}
\label{fig:inconsistency-farris}
\end{figure}

>>>>

Our key approach will be to decompose the likelihood as
\begin{equation}
\label{eq:partial-likelihood}
L^{(i)}(x', y', w') = H_{\tau'}(x,y;x',y',w') + L^{p,(i)}(x',y',w')
\end{equation}
where the first term is \eqref{eq:information} and the second term will be a weighted sum of the $\log$s of values from Table~\ref{tab:likelihoods}, the superscript indicating which choices for ancestral state likelihood values we choose.
Define
$$
\{\hat{x}, \hat{y}, \hat{w}, \hat{i}\} = \arg\max_{\{x,y,w,i\}} \ L^{p,(i)}(x,y,w).
$$
We have upper and lower bounds for \eqref{eq:partial-likelihood} in the form
$$
H_{\tau'}(x,y; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H_{\tau'}(x,y;x,y,y) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
$$

@@

We want to find values for $\{x,y\}$ where
$$
H_{fe}(x,y; 1, 1-c-2d, 0) + \hat{L}_{fe}(c,d) \ge H_{fa}(x,y; x, y, y) + \hat{L}_{fa}(a,b).
$$
Since this is a single inequality in two variables, we can plot it analytically; we need the region where
\begin{align*}
    0 &\ge H_{fa}(x,y; x, y, y) + \hat{L}_{fa}(a,b) - H_{fe}(x,y; 1, 1-c-2d, 0) - \hat{L}_{fe}(c,d) \\
      & = 2(12+2x^2+2y^2)\log((12+2x^2+2y^2)/8) \\
      &\qquad + 2(4-2x^2-2y^2)\log((4-2x^2-2y^2)/8)  + 8\log2 \\
      &\qquad + 2(1+x^2-y^2-x^2y^2)\log(1+x^2-y^2-x^2y^2) \\
      &\qquad + 2(1-x^2+y^2-x^2y^2)\log(1-x^2+y^2-x^2y^2) \\
      &\qquad + 2(1-x^2-y^2+x^2y^2)\log(1-x^2-y^2+x^2y^2) \\
      &\qquad + (1+x^2+y^2+4xy^2+x^2y^2)\log(1+x^2+y^2+4xy^2+x^2y^2) \\
      &\qquad + (1+x^2+y^2-4xy^2+x^2y^2)\log(1+x^2+y^2-4xy^2+x^2y^2) \\
      &\qquad - 16\log2 - 8(1+xy^2)\log(1+xy^2) - 8(1-xy^2)\log(1-xy^2) \\
      &\qquad - (2-4xy^2+2x^2y^2)\log(1-2xy^2+x^2y^4) \\
      &\qquad - (2+4xy^2+2x^2y^2)\log(1+2xy^2+x^2y^4) \\
      &\qquad - (4-4x^2y^2)\log(1-x^2y^4).
\end{align*}
The \texttt{sage} function in listing~\ref{list:sage} will generate Fig.~\ref{fig:inconsistency-farris}.

\begin{table}
    \begin{center}
\begin{verbatim}
var('x,y')
inconsistency = 2*(12+2*x^2+2*y^2)*log((12+2*x^2+2*y^2)/8) \
    + 2*(4-2*x^2-2*y^2)*log((4-2*x^2-2*y^2)/8) + 8*log(2.) \
    + 2*(1+x^2-y^2-x^2*y^2)*log(1+x^2-y^2-x^2*y^2) \
    + 2*(1-x^2+y^2-x^2*y^2)*log(1-x^2+y^2-x^2*y^2) \
    + 2*(1-x^2-y^2+x^2*y^2)*log(1-x^2-y^2+x^2*y^2) \
    + (1+x^2+y^2+4*x*y^2+x^2*y^2)*log(1+x^2+y^2+4*x*y^2+x^2*y^2) \
    + (1+x^2+y^2-4*x*y^2+x^2*y^2)*log(1+x^2+y^2-4*x*y^2+x^2*y^2) \
    - 16*log(2.) - 8*(1+x*y^2)*log(1+x*y^2) - 8*(1-x*y^2)*log(1-x*y^2) \
    - (2-4*x*y^2+2*x^2*y^2)*log(1-2*x*y^2+x^2*y^4) \
    - (2+4*x*y^2+2*x^2*y^2)*log(1+2*x*y^2+x^2*y^4) - (4-4*x^2*y^2)*log(1-x^2*y^4)

inconsistency_plot = implicit_plot(
    inconsistency(x,y)==0,
    (x,0,1),
    (y,0,1),
    axes_labels=['$x$', '$y$'],
    title='Region of inconsistency for Farris-generating topology',
    color='black'
)
\end{verbatim}
    \end{center}
\caption{Sage code to generate Fig.~\ref{fig:inconsistency-farris}}
\label{list:sage}
\end{table}

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{analytic-inconsistency}
\caption{Regions of inconsistency}
\label{fig:inconsistency-farris}
\end{figure}
